\section{Method}\label{sec:method}
\textcolor{red}{TODO: describe diffusion models as in between?}
\textcolor{red}{
    TODO: IMPORTANT! describe everything in a view that the model solves isomorphism OUTSIDE the model and not inside.
    make it a solution rather than a problem shift since all other methods solve iso in the model. write it in a way
    that sees the transfer as analogy between one-shot and sequential generators, ie, introducing a sequential algorithm class for
    exact decoding similar as for generative models.
}
\textcolor{red}{
    TODO: section should review not autoencoder but generators of graphs, where decoders are one special case.
}\\
\textcolor{red}{TODO: even GRALE is not 100\% unsupervised since labels are generated for matching}\\
\textcolor{red}{TODO: Molecule generators: this~\cite{mercadopractical}.}\\
\textcolor{red}{TODO: state that sequential generation is superior since isomorphism is a harder problem to solve than iterative decoding}\\
\textcolor{red}{explain that using iterative permutation invariant embeddings is some form of implicit marginalization of the permutation class}
\textcolor{red}{explain why to argue against direct reconstructions against a single random permutation}
\textcolor{red}{explain connection to GCPN for RL motivation and DGM as link prediction on the graph-level}
\textcolor{red}{explain that graph embeddings are a single source of truth for representation, other than adj. mat., PyG style or adjacency lists}

%GRAN does not even generate the graphs that are in their paper...
%https://github.com/lrjconan/GRAN/blob/master/runner/gran_runner.py#L320

good method must
\begin{itemize}
    \item be permutation-invariant
    \item handle-variable sized inputs (no padding)
    \item problematic: alignment methods are optimized towards one permutation, memoization instead of generalization
    \item avoid reintroducing node order at any point, graph level should be graph level and not mix
    \item any2graph and grale are not size-agnostic since they use a maximum size but do this with masking
    \item not solve GI internally
    \item have controlled generation for validity

\end{itemize}

In this section, we further analyze the existing body of work and identify desirable and undesirable properties
of deep graph decoders.

\paragraph{Addressing graph isomorphism.}
